---
title: "Bat Flight Presentation"
author: "Justin Benjamin & Renata Soljmosi"
date: "2024-04-08"
output:
  ioslides_presentation: default
  beamer_presentation: default
subtitle: Quantitative Methods in Ecology and Evolution
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

-   Big brown bats (*Eptesicus fuscus*)
-   Held in captivity at McMaster University
-   Previous work found they weren't willing/able to fly 
-   Planning a new study to examine whether exercise can help captive big browns fly/make them more willing to 

## Introduction continued

-   As this is an upcoming project, there are no data yet
-   Our project aims to simulate and analyze data 

## Hypothesis and predictions

- Hypothesis: Keeping bats active in captivity ensures better flight ability
- Predictions: Exercised bats will be more willing to fly and have longer flight time than non-exercised bats.


## Study design

-   30 bats measured for 60 consecutive days
-   Daily treatment of exercise or no exercise (control)
-   Record mass before and during 60 days
-   Flight tests every 3 days for every bat: Measure flight willingness as number of "shakes" and flight ability as flight time 

## Simulation process

-   Add proper code here
```{r cars, echo = TRUE}
summary(cars)
```

## Preliminary results

-   Add proper plot here
```{r pressure}
plot(pressure)
```


## Preliminary results continued

-   Add plot here


## Preliminary results cont


## Next steps

-   Running the simulation more times
-   Power analysis

## Questions?
