---
title: "Bat Flight Simulation"
author: "Justin Benjamin & Renata Soljmosi"
date: "April 8th 2024"
output:
  ioslides_presentation: default
  beamer_presentation: default
subtitle: Quantitative Methods in Ecology and Evolution
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

- Big brown bats (*Eptesicus fuscus*)
- Held in captivity at McMaster 
- Previously we found they weren't willing/able to fly in captivity
- Planning a new study to examine whether exercise via treadmill can help captive
bats become them more willing and able to fly
- Don't have data yet as this is an upcoming study
- Our project aims to simulate and analyze data 

## Study design

- Repeated measures study with 30 bats measured every 3 days over 60 days

- Predictor variables
    - Mass (grams)
    - Treatment (exercise or no exercise)
    - Sex (Most are females though)

- Response variables
    - Flight time
    - Number of "shakes" could be both a predictor for flight time and a response variable
    
## Hypothesis and predictions

- Hypothesis: Keeping bats active in captivity ensures better flight ability
- Predictions: 
    - Exercised bats will need less shakes to fly than non-exercised bats
    - Exercised bats will have longer flight times than non-exercised bats

## Simulation process

- Add proper code here
```{r cars, echo = TRUE}
summary(cars)
```

## Simulation process continued

## Preliminary results

- Add proper plot here
```{r pressure}
plot(pressure)
```

## Preliminary results continued

- Add plot here


## Preliminary results continued


## Next steps

- Running the simulation more times
- Simulation with shakes as count data (Negative binomial or Poisson models)

## Questions?
