---
title: "Bat Flight Presentation"
subtitle: "Quantitative Methods in Ecology and Evolution"
author: "Justin Benjamin & Renata Soljmosi"
date: "2024-04-08"
output:
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

-   Little brown bats (*Myotis lucifugus*)
-   Held in captivity at McMaster
-   Previous work found they couldn't/not willing to fly so this is a simulation of a new experiment to adress that.

## Introduction continued

- Introduction continued here

## Hypothesis and predictions

- Hypothesis: Keeping bats active in captivity ensures better flight ability
- Predictions: Exercised bats will be more willing to fly and have longer flight time than non-exercised bats.


## Study design

-   30 bats measured for 60 days
-   Mass before and during
-   Treatment of exercise or no exercise as a control
-   Measure flight time

## Simulation process

-   Add proper code here
```{r cars, echo = TRUE}
summary(cars)
```

## Preliminary results

-   Add proper plot here
```{r pressure}
plot(pressure)
```


## Preliminary results continued

-   Add plot here


## Preliminary results cont


## Next steps

-   Running the simulation more times
-   Power analysis

## Questions?
